Hijacking Simulators with Universal Probabilistic Programming

### 2019-08-01 paper/learning/notes.tex
minor additions to the text

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population.
To model a single population member quantum-based methods leveraging 
linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
the whole program, \emph{state}, in one iteration.
 Unfortunately, analytically writing the functional 
form of the state is, in this instance, impossible.   
The usability of current classical inference methods in this context is also problematic. 
However, by combining hijacking and amortizing parts of the simulator, in particular rejection sampling 
loops, we can not only correctly perform inference in such models, as the rejection sampling loops are 
now replaced with learnt functions, but we can also perform inference much more efficiently. 
This is because the rejection sampling loops are now replaced with learnt functions which enables us to correctly
construct the density and as they are amortized, evaluation is cheap and only has to be performed once per call.
This significantly reduces memory consumption and running time.

@@

However, to implement this in practice we are required to do the following:
 \begin{itemize}
    \item Create a sensible training procedure
    \item Have 2 additional functions implemented within pyprob
    \begin{enumerate} \item 1) That stores only the necessary raw samples from the desired rejection sampling addresses to be used for training.
                       \item 2) A function at \emph{test time} that replaces those given addresses with a learnt surrogate.
    \end{enumerate}
    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problems remain tractable.
  \end{itemize}
There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
currently not amenable to the current approach.

 \input{thesystem.tex}
\section{Limitations with different inference engines}

@@

\subsection{Sequential Monte Carlo}
Given multiple observations, we can ake the inference more computationally efficient 
by employing sequential monte carlo (SMC), although as we can only condition at the end 
of one full run of the simulation, computational speed increases are minimal.
SMC allows for conditioning on multiple observations, but the number of observations must
be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
could get dramatic speed increases over standard importance sampling.}

@@

In the standard setting we let $X$ be a random variable whose density is $p(x)$ that we cannot easily 
sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{p(x)}{Mg(x)}$, which we can do until acceptance. 
M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{p(x)}{g(x)}$ over the 
support of $X$. $M$ must also satify the following inequality $p(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $p(x)>0$. 
The standard algorithm is defined as follows:
\begin{algorithm}
  \small
   \caption{Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

As useful rejection sampling is, it has many short comings. 
In particular, to simulate proposals for certain distributions,
such as truncated normal distributions, regardless of the easy to sample from proposal
chosen the computational running time of such rejection statements can be shown to have an
expectation equal to infinity. Thus, the ability to learn and replace such rejection
sampling blocks with learnt surrogate functions, that are expressive enough to 
reciprocate the behavior of the rejection sampling block buys the user two things. 
A) It reduces the computational complexity of the simulator. 
B) As we now have a direct functional form of the density, we can correctly leverage
more computationally and sample efficient inference schemes. 
\section{Details}

@@

\section{ An overview of Amortized Inference schemes}
\textbf{Bradley: Todo: Tom has provided me with all the information to conduct this approach}
\begin{itemize}
  \item Whilst the variational posterior we choose as our proposal has a functional form 
  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
  which leads to problems when calculating the $KL$-divergence. 
  \item There are two approaches that we can take to calculating some divergence metric. 
  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from. 
  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
\end{itemize}

>>>>

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population, which presents challenges for current inference methods.
Because in many simulated settings we cannot correctly construct the functional form of the density of the 
desired program. 
This has spawned a large body of research in recent years to understand how to perform 
inference in likelihood free methods~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
for conditional density estimation. However, whilst many of these methods look to generate a surragote 
for the entire simulator, we propose amortizing only computational expensive sub-components 
of the simulator, which we can do in an automated way with probabilistic programming. 
This enables three things. 
1) It reduces the loss of information, as when simulators are amortized with in current
settings we require, potentially infinite samples, to provide a useful learnt representation of the simulator. 
2) It increases computational efficiency, whilst maintaining expressivity. 
3) It converts expensive rejection sampling loops into to learnt functions that can be 
easily sampled from and used to construct a functional form of the likelihood, that enables the 
target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, 
namely MCMC.

% To model a single population member quantum-based methods leveraging 
% linear superposition and entanglement provide an ability to explore multiple paths simultaneously maintaining 
% the whole program, \emph{state}, in one iteration.
%  Unfortunately, analytically writing the functional 
% form of the state is, in this instance, impossible.   
% 
In particular, we focus on combining the hijacking of simulators via probabilistic programming~\cite{baydin2018efficient} and 
state-of-the-art amortization schemes for density estimation \textbf{ Do home work on current papers}.
This enables such a process to happen in automatic way, simplifying existing set-ups, whilst providing
more expressive modelling capabilities. 
We are able to demonstrate this by replacing rejection sampling blocks, which are utilised within 
many prominent simulators to simulate densities that have no well defined cumulative density 
function, nor no known process to efficiently sample from such a density exists. 
By amortizing the rejection sampling blocks, during the inference stage when a simulator is run \emph{forword} we 
can replace such blocks with their leant surrogate, from which we can then construct the entire 
density of the simulator, which can then be utilized, \emph{correctly}, contrary to ~\cite{baydin2018efficient},
by sample efficient Markov chain Monte Carlo inference schemes, such as Hamiltonian Monte Carlo~(HMC),
random-walk Metropolis Hastings~(RMH), Light-weight Metropolis Hastings~(LMH) and many other schemes.
Whilst also significantly reducing memory consumption and computational running time of 
both the simulator and of the probabilistic programming system, as evaluating amortized functions 
is cheap and the reduction in rejection sampling blocks significantly reduces the number of 
unnecessary latent variables tracked~\cite{baydin2018efficient}.

@@

However, to construct this hybrid procedure requires both substantial engineering contributions 
and a description of the mathematical framework for these non-standard probabilistic programs. 
We provide both of these in this paper. As to implement this in practice we are required to do the following:
 \begin{itemize}
    \item An automated training procedure for any given simulator.
    \item Extend the protocols of existing probabilistic programming systems that hijack simulators, to enable:
    \begin{enumerate} \item 1) Only the necessary raw samples from the desired rejection sampling addresses to be collected for training.
                       \item 2) A function at inference time that replaces those given addresses with a learnt surrogate.
    \end{enumerate}
    \item Finer grained control over the actual latent variables of interest, to avoid unnecessary memory consumption and ensure that inference problems remain tractable.
  \end{itemize}

\textbf{There is no doubt that implementing some fo these features will be more invasive to the underlying simulator code than previously
just hijacking the raw-random number draws, but, it would enable us to attack difficulties in the inference procedures that are
currently not amenable to the current approach.}

\section{Limitations with different inference engines}

@@

\subsection{Sequential Monte Carlo}
Given multiple observations, we can ake the inference more computationally efficient 
by employing sequential monte carlo (SMC), although as we can only condition at the end 
of one full run of the simulation, computational speed increases are minimal.
SMC allows for conditioning on multiple observations, but the number of observations must
be fixed. \textbf{Note: If we could condition on observations within the simulator, we 
could get dramatic speed increases over standard importance sampling.We can actually do this 
in pyprob, but it requires knowing where to condition. That can be quite challenging 
for arbitrary code. }

@@

In the standard setting we let $X$ be a random variable whose density is $f(x)$ that we cannot easily 
sample from, but can evaluate. Using a random variable $Z$, whose density is $g(z)$  that we can efficiently
sample from, then we can generate a sample $X$ by sampling $Z$ and accept $Z$ with probability
$p_{accept} = \frac{f(x)}{Mg(x)}$, which we can do until acceptance. 
M is constant that has a finite bound, $M \in (1, \infty)$, on the likelihood ratio $\frac{f(x)}{g(x)}$ over the 
support of $X$. $M$ must also satify the following inequality $f(x) \leq M g(x)  \forall x$. 
This implies that the support of $Z$ must also contained the support of $X$, that is
$g(x) > 0$ whenever $f(x)>0$. 
The standard algorithm is defined as follows:
\begin{algorithm}
  \small
   \caption{Rejection Sampling}
    \begin{algorithmic}\small
       \State {At iteration $i$ s.t $i \geq 1$}
       \State $X_{i} \sim g(X_{i})$ 
       \State $U_{i} \sim \mathcal{U}(0,1)$
       \If{$U_{i} \leq M \frac{f(X_i)}{g(X_{i})}$}
       \State{ accept $X_i \sim f$}
       \Else
       \State{$i \gets i + 1$, repeat}
       \EndIf
\end{algorithmic}
\end{algorithm}

As useful rejection sampling is, it has many short comings. 
In particular, to simulate proposals for certain distributions,
such as truncated normal distributions, regardless of the easy to sample from proposal
chosen the computational running time of such rejection statements can be shown to have an
expectation equal to infinity. Thus, the ability to learn and replace such rejection
sampling blocks with learnt surrogate functions, that are expressive enough to 
reciprocate the behavior of the rejection sampling block buys the user two things. 
A) It reduces the computational complexity of the simulator. 
B) As we now have a direct functional form of the density, we can correctly leverage
more computationally and sample efficient inference schemes.

\section{Defining the Density and Sub-Densities within Simulators}

To construct the mathematical framework in this setting we employ the notation of~\cite{rainforth2017automating}, 
where \emph{raw samples} are represented by $ f_{a_i}(x_i | \eta)$ and can

In order to correctly characterize this problem we first provide a useful diagram demonstrating 
the density of the system.

 \input{thesystem.tex}

@@

\section{ An overview of Amortized Inference schemes}
\begin{itemize}
  \item Whilst the variational posterior we choose as our proposal has a functional form 
  and can be sampled, the functional form of the rejection sampling loops is not explicitly known,
  which leads to problems when calculating the $KL$-divergence. 
  \item There are two approaches that we can take to calculating some divergence metric. 
  \item 1) Using something like a non-parametric divergence estimator that makes use of Renyi and $L_{2}$ divergences (see \cite{basseville2013divergence} for details)
  \item 2) Use some form of kernel density to estimation to approximate the functional form of the rejection sampling 
  block, which can be used in place. From which we learn a variational approximation characterized by a trained network that we can sample from quickly. 
  \item An issue with approach 2) is that if I can approximate the density in functional form aren't there more efficient methods that we can use? (\textbf{Ask tom})
  \item Or, if assuming that the rejection sampling block can be arbitrary dimension, are we first learning a flow to represent that density and then learn a variational approximation of something that we can easily sample from. 
  \item But, this is very much the idea behind normalising flows, take a simple distribution that you can sample and then map it to the ``complicated'' density. 
  \item Stein Variational inference may be quite nice to use here~\cite{feng2017learning}, but it does make explicit assumptions in requiring the densities to be differentiable. 
  \item  For an overview of all variational methods, including the use of different divergence metrics see here ~\cite{8588399}
\end{itemize}

New questions to ask Tom

\begin{itemize}
  \item Isn't what we are doing just conditional density estimation? It seems
  like there is a lot of prior work here. I.e see this paper here for a good overview ~\cite{greenberg2019automatic}
  \item It does seem like this field has a large extensive past.
\end{itemize}
